---
title: "Cross Validation"
author: "Kennedy Wade"
date: "2025-11-11"
output: html_document
---



```{r}
library(tidyverse)
library(modelr)
library(mgcv)
```


```{r}
library(p8105.datasets)

set.seed(1)

```


# CV by hand
- The code chunk below imports data that is non-linear and shows increasing variance as the predictor increases. I like to use this setting because “model complexity” is easiest for me to understand when I can see it. However, “model complexity” is also an issue when you’re dealing with lots of predictors – you can’t “see” overfitting as easily, but it definitely happens.
```{r}
data("lidar")

lidar_df = 
  lidar |> 
  as_tibble() |>
  mutate(id = row_number())

lidar_df |> 
  ggplot(aes(x = range, y = logratio)) + 
  geom_point()
```



I’ll split this data into training and test sets (using anti_join!!), and replot showing the split. Our goal will be to use the training data (in black) to build candidate models, and then see how those models predict in the testing data (in red).
```{r}
train_df = sample_frac(lidar_df, size = .8)
test_df = anti_join(lidar_df, train_df, by = "id")

ggplot(train_df, aes(x = range, y = logratio)) + 
  geom_point() + 
  geom_point(data = test_df, color = "red")
```



I’ll fit three three models to the training data. Throughout, I’m going to use mgcv::gam for non-linear models – this is my go-to package for “additive models”,
```{r}
linear_mod = lm(logratio ~ range, data = train_df)
smooth_mod = mgcv::gam(logratio ~ s(range), data = train_df)
wiggly_mod = mgcv::gam(logratio ~ s(range, k = 30), sp = 10e-6, data = train_df)
```

to understand what these models have done, I'll plit the two gam fits 

```{r}
train_df |> 
  add_predictions(smooth_mod) |> 
  ggplot(aes(x = range, y = logratio)) + 
  geom_point() + 
  geom_line(aes(y = pred), color = "red")
```

```{r}
train_df |> 
  add_predictions(wiggly_mod) |> 
  ggplot(aes(x = range, y = logratio)) + 
  geom_point() + 
  geom_line(aes(y = pred), color = "red")
```

can also use the modelr::gather_predictions function – this is, essentially, a short way of adding predictions for several models to a data frame and then “pivoting” so the result is a tidy, “long” dataset that’s easily plottable.

```{r}
train_df |> 
  gather_predictions(linear_mod, smooth_mod, wiggly_mod) |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = range, y = logratio)) + 
  geom_point() + 
  geom_line(aes(y = pred), color = "red") + 
  facet_wrap(~model)
```


As a next step in my CV procedure, I’ll compute root mean squared errors (RMSEs) for each model.

```{r}
rmse(linear_mod, test_df)

rmse(smooth_mod, test_df)

rmse(wiggly_mod, test_df)

```


# CV using modelr

```{r}
cv_df = 
  crossv_mc(lidar_df, 100) 
```

****crossv_mc tries to be smart about memory – rather than repeating the dataset a bunch of times, it saves the data once and stores the indexes for each training / testing split using a resample object. This can be coerced to a dataframe, and can often be treated exactly like a dataframe. However, it’s not compatible with gam, so we have to convert each training and testing dataset (and lose that nice memory-saving stuff in the process) using the code below. It’s worth noting, though, that if all the models you want to fit use lm, you can skip this.

```{r}
cv_df |> pull(train) |> nth(1) |> as_tibble()

cv_df |> pull(test) |> nth(1) |> as_tibble()
```


```{r}
cv_df =
  cv_df |> 
  mutate(
    train = map(train, as_tibble),
    test = map(test, as_tibble))
```


I now have many training and testing datasets, and I’d like to fit my candidate models above and assess prediction accuracy as I did for the single training / testing split. To do this, I’ll fit models and obtain RMSEs using mutate + map & map2.
```{r}
cv_df = 
  cv_df |> 
  mutate(
    linear_mod  = map(train, \(df) lm(logratio ~ range, data = df)),
    smooth_mod  = map(train, \(df) gam(logratio ~ s(range), data = df)),
    wiggly_mod  = map(train, \(df) gam(logratio ~ s(range, k = 30), sp = 10e-6, data = df))) |> 
  mutate(
    rmse_linear = map2_dbl(linear_mod, test,rmse),
    rmse_smooth = map2_dbl(smooth_mod, test,rmse),
    rmse_wiggly = map2_dbl(wiggly_mod,test,rmse)
  )
```


Lets try try to look at this better
```{r}
cv_df |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_"
  ) |>
ggplot(aes(x = model, y = rmse)) + geom_violin()
```
















